Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2012-08-13 paper.tex
Added to paper.  Starting the empirical process proofs.  It's been moved to the appendix.

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data}
\author{Robby McKilliam, Andre Pollok, Bill Cowley}

@@

We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + X.
\]
The joint distribution of $R$ and $\Phi$ can be shown to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{\sqrt{r^2 - 2r\cos\phi + 1}}.
\]
The mean of $R e^{j\Phi}$ is equal to one because the mean of $X$ is zero.  So,
\[
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\]
where $\Re(\cdot)$ denotes the real part, and
\[
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\]
where $\Im(\cdot)$ denotes the imaginary part.  %The next lemma will be useful for the analysis of the least squares estimator.

@@

\begin{corollary}\label{cor:ampmse}
The mean square error of the amplitude estimator $\hat{\rho}$ is 
\[
\expect (\rho_0 - \hat{\rho})^2 = \frac{p B_1 + d B_2}{L} + \rho_0^2(1 - G(0))^2 + o(L^{-1/2}).
\] 
\end{corollary}
\begin{IEEEproof} Write
\begin{align*}
\expect (\rho_0 - \hat{\rho})^2 &= \expect \big( \rho_0(1 - G(0)) + G(0)\rho_0 - \hat{\rho} \big)^2 \\
&= \expect \big( \rho_0(1 - G(0)) - \hat{m}_L \big)^2 \\
&= \rho_0^2(1 - G(0))^2 + \expect \hat{m}_L^2 - 2\rho_0(1 - G(0))\expect\hat{m}_L.
\end{align*} 
From Theorem~\ref{thm:normality} it follows that $\expect \hat{m}_L^2 = \tfrac{B}{L} + o(L^{-1})$ and that $\expect \hat{m}_L = o(L^{-1/2})$. 
\end{IEEEproof}

@@

\begin{lemma}\label{lem:uniflawGGL}
The function $G_L$ converges almost surely and uniformly on $[-\pi, \pi)$ to the function $G$, that is,
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - G(\lambda)} \rightarrow 0
\] 
almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put $T(\lambda) = \expect G_L(\lambda)$ and write
\begin{align*}
&\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - G(\lambda)} \\
&\leq \sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T(\lambda) + T(\lambda) - G(\lambda)} \\
&\leq \sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T(\lambda)} +  \sup_{\lambda \in [-\pi, \pi)}\sabs{T(\lambda) - G(\lambda)}.
\end{align*}
Now,
\begin{align*}
T(\lambda) - G(\lambda) &= \big(\tfrac{\abs{P}}{L} - p\big)h_1(\lambda) + \big(\tfrac{\abs{D}}{L} - d\big)h_2(\lambda) \\
&= o(1) h_1(\lambda) + o(1) h_2(\lambda)
\end{align*}
where $o(1)$ denotes a number converging to zero as $L \rightarrow \infty$.  Since 
\[
h_1(\lambda) = \expect R_1\cos(\lambda + \Phi_1) \leq \expect R_1,
\]
and 
\[
h_2(\lambda) = \expect R_1\cos\fracpart{\lambda + \Phi_1} \leq \expect R_1
\] 
for all $\lambda \in [-\pi, \pi)$, it follows that 
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{T(\lambda) - G(\lambda)} \leq o(1) \expect R_1 \rightarrow 0
\]  
as $L\rightarrow \infty$.  Lemma~\ref{lem:uniflawTGL} shows that 
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T(\lambda)} \rightarrow 0
\]
almost surely as $L\rightarrow \infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:uniflawTGL} 
Put $T(\lambda) = \expect G_L(\lambda)$, then
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T(\lambda)} \rightarrow 0
\] 
almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put $D_L(\lambda) = G_L(\lambda) - T(\lambda)$ and let
\[
\lambda_n = \tfrac{2\pi}{N}(n-1) - \pi, \qquad n = 1, \dots, N
\]
be $N$ points uniformly spaced on the interval $[-\pi, \pi)$.  Let $L_n$ be the interval $[\lambda_n, \lambda_n + \tfrac{2\pi}{N})$ and observe that $L_1, \dots, L_N$ partition $[-\pi, \pi)$.  Now
\begin{align*}
\sup_{\lambda \in [-\pi, \pi)}&\sabs{D_L(\lambda)} \\
&= \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n) + D_L(\lambda_n)} \\
&\leq U_L + V_L,
\end{align*}
where 
\[
U_L = \sup_{n = 1,\dots,N}\sabs{D_L(\lambda_n)},
\]
\[
V_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n)}.
\]
Lemma~\ref{lem:supDLlambdan} shows that for any finite $N$, and any $\epsilon > 0$, 
\[
\prob\left\{ \lim_{L\rightarrow\infty} U_L > \epsilon \right\} = 0,
\]
i.e., $U_L \rightarrow 0$ almost surely as $L\rightarrow\infty$.  Lemma~\ref{lem:supsupDLlambda} shows that for any $\epsilon > 0$,
\[
\prob\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} = 0.
\] 
Choose $N$ large enough that $4\pi\expect R_i < \epsilon N$ and
\begin{align*}
\prob& \left\{ \lim_{L\rightarrow\infty} \sup_{\lambda \in [-\pi, \pi)}\sabs{D_L(\lambda)} > 3\epsilon \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty}(U_L + V_L) > 3\epsilon \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty} (U_L + V_L) > \epsilon + \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty} U_L > \epsilon\right\} +  \prob\left\{ \lim_{L\rightarrow\infty} V_L > \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} \\
&= 0.
\end{align*}
Thus, $\sup_{\lambda \in [-\pi, \pi)}\sabs{D_L(\lambda)} \rightarrow 0$ almost surely as $L\rightarrow\infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:supDLlambdan}
If $N$ is finite then $U_L \rightarrow 0$ almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put
\[
Z_i(\lambda) = \begin{cases}
R_i\cos(\lambda + \Phi_i) , & i \in P \\
R_i\cos\sfracpart{\lambda + \Phi_i} , & i \in D
\end{cases}
\]
so that
\[
D_L(\lambda) = G_L(\lambda) - T(\lambda) = \frac{1}{L}\sum_{i=1}^{L} \big( Z_i(\lambda) - \expect Z_i(\lambda) \big).
\]
Now $Z_1(\lambda_n), \dots,Z_L(\lambda_n)$ are independent with finite variance (because $\expect R_i^2$ is finite), so for each $n =1, \dots, N$,
\[
\abs{D_L(\lambda_n)} = \abs{\frac{1}{L}\sum_{i=1}^{L} \big( Z_i(\lambda_n) - \expect Z_i(\lambda)\big)} \rightarrow 0
\]
almost surely as $L\rightarrow\infty$ by the Kolmogorov's strong law of large numbers~\cite[Theorem 2.3.10]{SenSinger_large_sample_stats_1993}.  Now, since $N$ is finite,
\[
U_L = \sup_{n=1,\dots,N}\sabs{D_L(\lambda_n)} \leq \sum_{n=1}^N \sabs{D_L(\lambda_n)} \rightarrow 0
\]
almost surely to zero as $L \rightarrow \infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:supsupDLlambda} For any $\epsilon > 0$,
\[
\prob\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{2\pi}{N}\expect R_i \right\} = 0.
\]
\end{lemma}
\begin{IEEEproof}
Observe that
\begin{align*}
&\sabs{D_L(\lambda) - D_L(\lambda_n)} \\
&=  \abs{G_L(\lambda) - T(\lambda) - G_L(\lambda_n) + T(\lambda_n)} \\
&\leq  \abs{G_L(\lambda) - G_L(\lambda_n)} + \abs{\expect G_L(\lambda) - \expect G_L(\lambda_n)} \\
&\leq \abs{G_L(\lambda) - G_L(\lambda_n) } + \expect \abs{G_L(\lambda) - G_L(\lambda_n)},
\end{align*}
the last line following from Jensens inequality.  Put
\[
C_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \abs{G_L(\lambda) - G_L(\lambda_n) },
\]
so that
\[
V_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n)} \leq C_L + \expect C_L,
\]
because 
\[
 \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \expect \abs{G_L(\lambda) - G_L(\lambda_n)} \leq \expect C_L
\]
follows by moving expectation outside the supremum.  Lemma~\ref{lem:CL} shows that $\expect C_L \leq \tfrac{2\pi}{N}\expect R_1$ and also that
\[
\prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\]
So,
\begin{align*}
\prob&\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} (C_L + \expect C_L) > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\end{align*}
\end{IEEEproof}

@@

We first prove the asymptotic normality of $\sqrt{L} \hat{\lambda}_L$.  Once this is done we will be able to prove the normality of $\sqrt{L} m_L$.  Recall that $\hat{\lambda}_L$ is the maximiser of the function $G_L$ defined in~\eqref{eq:GLdefn}.  The proof is complicated by the fact that $G_L$ is not differentiable everywhere due to the function $\fracpart{\cdot}$ not being differentiable at multiples of $\tfrac{\pi}{M}$.  This prevents the use of ``standard approaches'' to proving normality that are based on the mean value theorem~\cite{vonMises_diff_stats_1947,vanDerVart1971_asymptotic_stats,Pollard_new_ways_clts_1986,Pollard_conv_stat_proc_1984,Pollard_asymp_empi_proc_1989}.  However, we show in Lemma~\ref{lem:diffatlambdaL} that the derivative $G_L^\prime$ does exist, and is equal to zero, at $\hat{\lambda}_L$. Define the function
\[
R_L(\lambda) = \frac{1}{L} \sum_{i \in P} R_i \sin(\lambda + \Phi_i) + \frac{1}{L} \sum_{i \in D} R_i \sin\sfracpart{\lambda + \Phi_i}.
\]
Whenever $G_L(\lambda)$ is differentiable $G_L^\prime(\lambda) = R_L(\lambda)$ so $R_L(\hat{\lambda}_L) = G_L^\prime(\hat{\lambda}_L) = 0$ by Lemma~\ref{lem:diffatlambdaL}.  Let $Q_L(\lambda) = \expect R_L(\lambda) - \expect R_L(0)$ and write
\begin{align*}
0 &= R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) + Q_L(\hat{\lambda}_L) \\
&= \sqrt{L}\big( R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) \big) + \sqrt{L}Q_L(\hat{\lambda}_L).
\end{align*}
Lemma~\ref{lem:Qconv} shows that
\[
\sqrt{L} Q_L(\hat{\lambda}_L) = \sqrt{L} \hat{\lambda}_L\big( p + Hd  + o_P(1) \big)
\]
where $o_P(1)$ denotes a number converging in probablity to zero as $L \rightarrow \infty$, and the constants $p,d$ and $H$ are defined in the statement of Theorems~\ref{thm:consistency}~and~\ref{thm:normality}.  Lemma~\ref{lem:empiricprocc} shows that
\[
\sqrt{L}\big( R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} R_L(0).
\]
It follows from the three equations above that,
\[
0 = o_P(1) + \sqrt{L}R_L(0) + \sqrt{L} \hat{\lambda}_L \big( p + Hd  + o_P(1) \big)
\]
and rearranging gives,
\[
\sqrt{L} \hat{\lambda}_L = o_P(1) - \frac{\sqrt{L}R_L(0)}{p + Hd  + o_P(1)}.
\]
Lemma~\ref{lem:convdistGLdash} shows that the distribution of $\sqrt{L}R_L(0)$ converges to the normal with zero mean and variance $pA_1 + dA_2$ where $A_1$ and $A_2$ are real constants defined in the statement of Thereom~\ref{thm:normality}.  It follows that the distribution of $\sqrt{L}\hat{\lambda}_L$ converges to the normal with zero mean and variance
\[
\frac{pA_1 + dA_2}{(p + Hd)^2}.
\]

We now analyse the asymptotic distribution of $\sqrt{L} m_L$.  Let $T(\lambda) = \expect G_L(\lambda)$.  Using~\eqref{eq:rhoGLZ},
\begin{align*}
\sqrt{L} m_L &= \sqrt{L} \rho_0 \big( G_L(\hat{\lambda}_L) - G(0) \big) \\
&= \sqrt{L} \rho_0 \big( G_L(\hat{\lambda}_L) - T(\hat{\lambda}_L) + T(\hat{\lambda}_L) - G(0) \big).
\end{align*}
Lemma~\ref{lem:empiricprocforrho} shows that 
\[
\sqrt{L}\big( G_L(\hat{\lambda}_L) - T(\hat{\lambda}_L)  \big) = o_P(1) + X_L,
\]
where $X_L = \sqrt{L}\big( G_L(0) - T(0)  \big)$.  Lemma~\ref{lem:HLtoG} shows that
\[
%\sqrt{L}\big( T(\hat{\lambda}_L) - G(0) \big) = \sqrt{L} \hat{\lambda}_L \big( ? + o_P(1) \big).
\sqrt{L}\big( T(\hat{\lambda}_L) - G(0) \big) = o_P(1).
\]
It follows that
\[
%\sqrt{L} m_L = o_P(1) + X_L  + \hat{\lambda}_L \big( ? + o_P(1) \big).
\sqrt{L} m_L =  X_L + o_P(1). 
\]
Lemma~\ref{lem:XL} shows that the distrubtion of $X_L$ converges to the normal with zero mean and variance
\[
 p B_1 + d B_2 
\]
where $B_1$ and $B_2$ are defined in the statement of Theorem~\ref{thm:normality}.  As $X_L$ does not depend on $\hat{\lambda}_L$, it follows that $\covar(X_L, \sqrt{L}\hat{\lambda}_L) = 0$.  Thus,
\[
\covar(\sqrt{L}\hat{m}_L, \sqrt{L}\hat{\lambda}_L) = \covar(X_L + o_P(1), \sqrt{L}\hat{\lambda}_L) \rightarrow 0
\]
as $L \rightarrow \infty$.  It remains to prove the lemmas that we have used.

>>>>

%opening
\title{Carrier phase and amplitude estimation for phase shift keying using pilots and data}
\author{Robby McKilliam, Andre Pollok, Bill Cowley 
\thanks{
Supported under the Australian Government’s Australian Space Research Program.
Robby McKilliam, Andre Pollok and Bill Cowley are with the Institute for Telecommunications Research, The University of South Australia, SA, 5095.
}}

@@

We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying, 
\[
R e^{j\Phi} = 1 + X.
\]
The joint distribution of $R$ and $\Phi$ can be shown to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{\sqrt{r^2 - 2r\cos\phi + 1}}.
\]
The mean of $R e^{j\Phi}$ is equal to one because the mean of $X$ is zero.  So,
\begin{equation}\label{eq:expectRepartRphi}
\expect \Re(R e^{j\Phi}) = \expect R \cos(\Phi) = 1,
\end{equation}
where $\Re(\cdot)$ denotes the real part, and
\begin{equation}\label{eq:expectImpartRphi}
\expect \Im(R e^{j\Phi}) = \expect R \sin(\Phi) = 0,
\end{equation}
where $\Im(\cdot)$ denotes the imaginary part.  %The next lemma will be useful for the analysis of the least squares estimator.

@@

\begin{corollary}\label{cor:ampmse}
The mean square error of the amplitude estimator $\hat{\rho}$ is 
\[
\expect (\rho_0 - \hat{\rho})^2 = \rho_0^2\frac{p B_1 + d B_2}{L} + \rho_0^2(1 - G(0))^2 + o(L^{-1/2}).
\] 
\end{corollary}
\begin{IEEEproof} Write
\begin{align*}
\expect (\rho_0 - \hat{\rho})^2 &= \expect \big( \rho_0(1 - G(0)) + G(0)\rho_0 - \hat{\rho} \big)^2 \\
&= \expect \big( \rho_0(1 - G(0)) - \hat{m}_L \big)^2 \\
&= \rho_0^2(1 - G(0))^2 + \expect \hat{m}_L^2 - 2\rho_0(1 - G(0))\expect\hat{m}_L.
\end{align*} 
From Theorem~\ref{thm:normality} it follows that $\expect \hat{m}_L^2 = \tfrac{B}{L} + o(L^{-1})$ and that $\expect \hat{m}_L = o(L^{-1/2})$. 
\end{IEEEproof}

@@

\begin{lemma}\label{lem:uniflawGGL}
The function $G_L$ converges almost surely and uniformly on $[-\pi, \pi)$ to the function $G$, that is,
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - G(\lambda)} \rightarrow 0
\] 
almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put $T_L(\lambda) = \expect G_L(\lambda)$ and write
\begin{align*}
&\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - G(\lambda)} \\
&\leq \sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T_L(\lambda) + T_L(\lambda) - G(\lambda)} \\
&\leq \sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T_L(\lambda)} +  \sup_{\lambda \in [-\pi, \pi)}\sabs{T_L(\lambda) - G(\lambda)}.
\end{align*}
Now,
\begin{align*}
T_L(\lambda) - G(\lambda) &= \big(\tfrac{\abs{P}}{L} - p\big)h_1(\lambda) + \big(\tfrac{\abs{D}}{L} - d\big)h_2(\lambda) \\
&= o(1) h_1(\lambda) + o(1) h_2(\lambda)
\end{align*}
where $o(1)$ denotes a number converging to zero as $L \rightarrow \infty$.  Since 
\[
h_1(\lambda) = \expect R_1\cos(\lambda + \Phi_1) \leq \expect R_1,
\]
and 
\[
h_2(\lambda) = \expect R_1\cos\fracpart{\lambda + \Phi_1} \leq \expect R_1
\] 
for all $\lambda \in [-\pi, \pi)$, it follows that 
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{T_L(\lambda) - G(\lambda)} \leq o(1) \expect R_1 \rightarrow 0
\]  
as $L\rightarrow \infty$.  Lemma~\ref{lem:uniflawTGL} shows that 
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T_L(\lambda)} \rightarrow 0
\]
almost surely as $L\rightarrow \infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:uniflawTGL} 
Put $T_L(\lambda) = \expect G_L(\lambda)$, then
\[
\sup_{\lambda \in [-\pi, \pi)}\sabs{G_L(\lambda) - T_L(\lambda)} \rightarrow 0
\] 
almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put $D_L(\lambda) = G_L(\lambda) - T_L(\lambda)$ and let
\[
\lambda_n = \tfrac{2\pi}{N}(n-1) - \pi, \qquad n = 1, \dots, N
\]
be $N$ points uniformly spaced on the interval $[-\pi, \pi)$.  Let $L_n$ be the interval $[\lambda_n, \lambda_n + \tfrac{2\pi}{N})$ and observe that $L_1, \dots, L_N$ partition $[-\pi, \pi)$.  Now
\begin{align*}
\sup_{\lambda \in [-\pi, \pi)}&\sabs{D_L(\lambda)} \\
&= \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n) + D_L(\lambda_n)} \\
&\leq U_L + V_L,
\end{align*}
where 
\[
U_L = \sup_{n = 1,\dots,N}\sabs{D_L(\lambda_n)},
\]
\[
V_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n)}.
\]
Lemma~\ref{lem:supDLlambdan} shows that for any finite $N$, and any $\epsilon > 0$, 
\[
\prob\left\{ \lim_{L\rightarrow\infty} U_L > \epsilon \right\} = 0,
\]
i.e., $U_L \rightarrow 0$ almost surely as $L\rightarrow\infty$.  Lemma~\ref{lem:supsupDLlambda} shows that for any $\epsilon > 0$,
\[
\prob\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} = 0.
\] 
Choose $N$ large enough that $4\pi\expect R_i < \epsilon N$ and
\begin{align*}
\prob& \left\{ \lim_{L\rightarrow\infty} \sup_{\lambda \in [-\pi, \pi)}\sabs{D_L(\lambda)} > 3\epsilon \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty}(U_L + V_L) > 3\epsilon \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty} (U_L + V_L) > \epsilon + \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} \\
&\leq \prob\left\{\lim_{L\rightarrow\infty} U_L > \epsilon\right\} +  \prob\left\{ \lim_{L\rightarrow\infty} V_L > \epsilon + \tfrac{4\pi }{N}\expect R_i \right\} \\
&= 0.
\end{align*}
Thus, $\sup_{\lambda \in [-\pi, \pi)}\sabs{D_L(\lambda)} \rightarrow 0$ almost surely as $L\rightarrow\infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:supDLlambdan}
If $N$ is finite then $U_L \rightarrow 0$ almost surely as $L \rightarrow \infty$.
\end{lemma}
\begin{IEEEproof}
Put
\[
Z_i(\lambda) = \begin{cases}
R_i\cos(\lambda + \Phi_i) , & i \in P \\
R_i\cos\sfracpart{\lambda + \Phi_i} , & i \in D
\end{cases}
\]
so that
\[
D_L(\lambda) = G_L(\lambda) - T_L(\lambda) = \frac{1}{L}\sum_{i=1}^{L} \big( Z_i(\lambda) - \expect Z_i(\lambda) \big).
\]
Now $Z_1(\lambda_n), \dots,Z_L(\lambda_n)$ are independent with finite variance (because $\expect R_i^2$ is finite), so for each $n =1, \dots, N$,
\[
\abs{D_L(\lambda_n)} = \abs{\frac{1}{L}\sum_{i=1}^{L} \big( Z_i(\lambda_n) - \expect Z_i(\lambda)\big)} \rightarrow 0
\]
almost surely as $L\rightarrow\infty$ by the Kolmogorov's strong law of large numbers~\cite[Theorem 2.3.10]{SenSinger_large_sample_stats_1993}.  Now, since $N$ is finite,
\[
U_L = \sup_{n=1,\dots,N}\sabs{D_L(\lambda_n)} \leq \sum_{n=1}^N \sabs{D_L(\lambda_n)} \rightarrow 0
\]
almost surely to zero as $L \rightarrow \infty$.
\end{IEEEproof}

\begin{lemma}\label{lem:supsupDLlambda} For any $\epsilon > 0$,
\[
\prob\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{2\pi}{N}\expect R_i \right\} = 0.
\]
\end{lemma}
\begin{IEEEproof}
Observe that
\begin{align*}
&\sabs{D_L(\lambda) - D_L(\lambda_n)} \\
&=  \abs{G_L(\lambda) - T_L(\lambda) - G_L(\lambda_n) + T_L(\lambda_n)} \\
&\leq  \abs{G_L(\lambda) - G_L(\lambda_n)} + \abs{\expect G_L(\lambda) - \expect G_L(\lambda_n)} \\
&\leq \abs{G_L(\lambda) - G_L(\lambda_n) } + \expect \abs{G_L(\lambda) - G_L(\lambda_n)},
\end{align*}
the last line following from Jensens inequality.  Put
\[
C_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \abs{G_L(\lambda) - G_L(\lambda_n) },
\]
so that
\[
V_L = \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n}\sabs{D_L(\lambda) - D_L(\lambda_n)} \leq C_L + \expect C_L,
\]
because 
\[
 \sup_{n = 1,\dots,N}\sup_{\lambda \in L_n} \expect \abs{G_L(\lambda) - G_L(\lambda_n)} \leq \expect C_L
\]
follows by moving expectation outside the supremum.  Lemma~\ref{lem:CL} shows that $\expect C_L \leq \tfrac{2\pi}{N}\expect R_1$ and also that
\[
\prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\]
So,
\begin{align*}
\prob&\left\{ \lim_{L \rightarrow \infty} V_L > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} (C_L + \expect C_L) > \epsilon + \tfrac{4\pi}{N}\expect R_1 \right\} \\
&\leq \prob\left\{ \lim_{L \rightarrow \infty} C_L > \epsilon + \tfrac{2\pi}{N}\expect R_1 \right\} = 0.
\end{align*}
\end{IEEEproof}

@@

We first prove the asymptotic normality of $\sqrt{L} \hat{\lambda}_L$.  Once this is done we will be able to prove the normality of $\sqrt{L} m_L$.  Recall that $\hat{\lambda}_L$ is the maximiser of the function $G_L$ defined in~\eqref{eq:GLdefn}.  The proof is complicated by the fact that $G_L$ is not differentiable everywhere due to the function $\fracpart{\cdot}$ not being differentiable at multiples of $\tfrac{\pi}{M}$.  This prevents the use of ``standard approaches'' to proving normality that are based on the mean value theorem~\cite{vonMises_diff_stats_1947,vanDerVart1971_asymptotic_stats,Pollard_new_ways_clts_1986,Pollard_conv_stat_proc_1984,Pollard_asymp_empi_proc_1989}.  However, we show in Lemma~\ref{lem:diffatlambdaL} that the derivative $G_L^\prime$ does exist, and is equal to zero, at $\hat{\lambda}_L$. Define the function
\begin{equation}\label{eq:RLdef}
R_L(\lambda) = \frac{1}{L} \sum_{i \in P} R_i \sin(\lambda + \Phi_i) + \frac{1}{L} \sum_{i \in D} R_i \sin\sfracpart{\lambda + \Phi_i}.
\end{equation}
Whenever $G_L(\lambda)$ is differentiable $G_L^\prime(\lambda) = R_L(\lambda)$ so $R_L(\hat{\lambda}_L) = G_L^\prime(\hat{\lambda}_L) = 0$ by Lemma~\ref{lem:diffatlambdaL}.  Let $Q_L(\lambda) = \expect R_L(\lambda) - \expect R_L(0)$ and write
\begin{align*}
0 &= R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) + Q_L(\hat{\lambda}_L) \\
&= \sqrt{L}\big( R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) \big) + \sqrt{L}Q_L(\hat{\lambda}_L).
\end{align*}
Lemma~\ref{lem:Qconv} shows that
\[
\sqrt{L} Q_L(\hat{\lambda}_L) = \sqrt{L} \hat{\lambda}_L\big( p + Hd  + o_P(1) \big)
\]
where $o_P(1)$ denotes a number converging in probablity to zero as $L \rightarrow \infty$, and the constants $p,d$ and $H$ are defined in the statement of Theorems~\ref{thm:consistency}~and~\ref{thm:normality}.  Lemma~\ref{lem:empiricprocc} shows that
\[
\sqrt{L}\big( R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} R_L(0).
\]
It follows from the three equations above that,
\[
0 = o_P(1) + \sqrt{L}R_L(0) + \sqrt{L} \hat{\lambda}_L \big( p + Hd  + o_P(1) \big)
\]
and rearranging gives,
\[
\sqrt{L} \hat{\lambda}_L = o_P(1) - \frac{\sqrt{L}R_L(0)}{p + Hd  + o_P(1)}.
\]
Lemma~\ref{lem:convdistGLdash} shows that the distribution of $\sqrt{L}R_L(0)$ converges to the normal with zero mean and variance $pA_1 + dA_2$ where $A_1$ and $A_2$ are real constants defined in the statement of Thereom~\ref{thm:normality}.  It follows that the distribution of $\sqrt{L}\hat{\lambda}_L$ converges to the normal with zero mean and variance
\[
\frac{pA_1 + dA_2}{(p + Hd)^2}.
\]

We now analyse the asymptotic distribution of $\sqrt{L} m_L$.  Let $T_L(\lambda) = \expect G_L(\lambda)$.  Using~\eqref{eq:rhoGLZ},
\begin{align*}
\sqrt{L} m_L &= \sqrt{L} \rho_0 \big( G_L(\hat{\lambda}_L) - G(0) \big) \\
&= \sqrt{L} \rho_0 \big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L) + T_L(\hat{\lambda}_L) - G(0) \big).
\end{align*}
Lemma~\ref{lem:empiricprocforrho} shows that 
\[
\sqrt{L}\big( G_L(\hat{\lambda}_L) - T_L(\hat{\lambda}_L)  \big) = o_P(1) + X_L,
\]
where $X_L = \sqrt{L}\big( G_L(0) - T_L(0)  \big)$.  Lemma~\ref{lem:HLtoG} shows that
\[
%\sqrt{L}\big( T_L(\hat{\lambda}_L) - G(0) \big) = \sqrt{L} \hat{\lambda}_L \big( ? + o_P(1) \big).
\sqrt{L}\big( T_L(\hat{\lambda}_L) - G(0) \big) = o_P(1).
\]
It follows that
\[
%\sqrt{L} m_L = o_P(1) + X_L  + \hat{\lambda}_L \big( ? + o_P(1) \big).
\sqrt{L} m_L =  \rho_0 X_L + o_P(1). 
\]
Lemma~\ref{lem:XL} shows that the distrubtion of $X_L$ converges to the normal with zero mean and variance $p B_1 + d B_2$ as $L\rightarrow\infty$ where $B_1$ and $B_2$ are defined in the statement of Theorem~\ref{thm:normality}.  Thus, the distribution of $\sqrt{L} m_L$ converges to the normal with zero mean and variance $\rho_0^2(p B_1 + d B_2)$ as required.  Because $X_L$ does not depend on $\hat{\lambda}_L$, it follows that $\covar(X_L, \sqrt{L}\hat{\lambda}_L) = 0$, and so,
\[
\covar(\sqrt{L}\hat{m}_L, \sqrt{L}\hat{\lambda}_L) = \covar(\rho_0 X_L + o_P(1), \sqrt{L}\hat{\lambda}_L) \rightarrow 0
\]
as $L \rightarrow \infty$.  It remains to prove the lemmas that we have used.
